---
title: "Using the TcGSA package"
author: "Anthony Devaux"
date: "`r Sys.Date()`"
output: 
  html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
  pdf_document:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Using TcGSA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r pre, echo=FALSE, warning=FALSE}
library(knitr)
```

# Overview of TcGSA

Le package Time-course Gene Set Analysis (*Hejblum et al., 2012*) utilise la méthode des Gene Sets Analysis qui consiste au regroupement de groupes de genes en fonction de leur utilité biologique. Ce package permet l'étude de données longitudinales afin de détecter si l'expression génique au sein des groupes de gène evolue avec le temps. Ce package se base sur les modèles linéaires mixtes afin de prendre en compte les effets aléatoires entre les individus et entre les patients.

# Getting started using TcGSA


## Load package

```{r library, eval=FALSE}
library(TcGSA)
```

## Data structures

### Gene set object

Les groupes de gènes sont définis par une matrice où on y retrouve pour chaque gene set la liste de tous les gènes associés. Les groupes de gènes sont constitués à partir des fonctions biologiques communes aux gènes.

### Gene expression matrix

This matrix contains the gene expression (in cells) for each gene (in row) of each sample of patients (in column). You can see below an example (called `expr_gen`) in this matrix.

```{r expr_gen_1, echo=FALSE}
expr_gen <- matrix(data = c(4.967101,5.215882,4.565933,4.402265,
                            4.398911,4.744463,4.578123,4.473504,
                            4.384093,5.038972,4.696965,4.598726,
                            4.927489,5.209726,4.306206,4.589389),
                   nrow = 4, ncol = 4)
rownames(expr_gen) <- c("ILMN_1762337","ILMN_2055271","ILMN_2383229",
                        "ILMN_1806310")
colnames(expr_gen) <- c("X6116733008_L","X6116733022_E","X6116733029_A",
                        "X6116733068_L")
```
```{r expr_gen_2}
expr_gen
```

### Experimental variables matrix

The experimental variables matrix contains for each sample (in row), several variables (in column). Required for the matrix :

* `subject_name` to identify the patients 
* `time_func` to identify the time measurements
* __For more than one treatment group :__ `group_name` to identify the group of patients

See below an example of matrix (called `design_gen`) with experimental variables

```{r design_gen_1, echo=FALSE}
design_gen <- matrix(data = c("X6116733008_L","X6116733022_E","X6116733029_A",
                        "X6116733068_L",
                        "P001","P002","P001","P002",
                        "Vaccine","Saline","Vaccine","Saline",
                        0,12,0,12),
                   nrow = 4, ncol = 4)
colnames(design_gen) <- c("Sample_name","Patient_ID","Group",
                        "TimePoint")
```
```{r design_gen_2}
design_gen
```



# How to use TcGSA for one treatment group ? (Data from DALIA-1 trial)

## Likelihood ratios test

This function provides the result of likelihood ratio test for each gene set.

```{r LR_ST, eval=FALSE}
tcgsa_result <- TcGSA.LR(expr = expr_gen, 
         gmt = , 
         design = design_gen, 
         subject_name = "Patient_ID", 
         time_name = "TimePoint", 
         time_func = "linear", 
         crossedRandom = TRUE)
```

Now `tcgsa_result` is your TcGSA object.

`signifLRT.TcGSA` returns severals informations such as the significant gene sets and their ajusted p-values with defaut options Benjamini-Yekutieli correction and 5% threshold.

```{r signifLRT_ST_1, echo=FALSE}
signifLRT <- matrix(data = c("M1.1","M3.2","M3.5","M3.6",
                           0.000000e+00,0.000000e+00,6.973542e-03,1.320619e-02,
                             "Platelets","Inflammation","Cell Cycle","Cytotoxic/NK Cell"),
                   nrow = 4, ncol = 3)
rownames(signifLRT) <- c("1","2","3","4")
colnames(signifLRT) <- c("GeneSet","AdjPval","desc")
```

```{r signifLRT_ST_2, eval=FALSE}
signifLRT.TcGSA(tcgsa_result)$mixedLRTadjRes
```
```{r signifLRT_ST_3, echo=FALSE}
signifLRT
```


## Plot the median of standardized gene expression

```{r plot1GS_ST, eval=FALSE}
plot1GS(expr = , 
        gmt = , 
        Subject_ID = , 
        TimePoint = , 
        indiv = , 
        clustering = , 
        time_unit = , 
        geneset.name = , 
        title="",
        margins=0.4,
        lab.cex=0.37,
        axis.cex=0.37,
        line.size=0.45,
        gg.add=list(theme(legend.position="none"),
                    ylim(-1.26,1.26)
        ) )
```


# How to use TcGSA for several treatment group ? (Example from *Obermoser et al., 2013* data)

## Likelihood ratios test

This function provides the result of likelihood ratio test for each gene set.

```{r LR_MT, eval=FALSE}
tcgsa_result <- TcGSA.LR(expr = expr_gen, 
         gmt = , 
         design = design_gen, 
         subject_name = "Patient_ID", 
         time_name = "TimePoint", 
         time_func = "linear", 
         crossedRandom = TRUE,
         group_name = "group")
```

## Plot clusters of significant gene sets

### Make clusters from tcgsa object

```{r clust_MT, eval=FALSE}
clust <- clustTrend(tcgs = tcgsa_result, 
                    expr=tcgsa_result$Estimations,
                    Subject_ID=design_gen$Patient_ID,
                    TimePoint=design_gen$TimePoint,
                    threshold=0.1,
                    group_of_interest="vaccine",
                    max_trends=2, B=50)
```

### Plot heatmap of clusters of significant gene sets

```{r heatmap_MT, eval=FALSE}
plot(x=tcgsa_result, expr=tcgsa_result$Estimations,
     Subject_ID=design_gen$Patient_ID,
     TimePoint=design_gen$TimePoint,
     group_of_interest="vaccine",
     clust_trends=clust,
     legend.breaks=seq(from=-2,to=2, by=0.01), time_unit="M",
     subtitle="Vaccine vs Saline", cex.label.row=2.5, cex.label.col=3,
     heatmap.width=0.4, dendrogram.size=0.25, margins=c(5,23),
     heatKey.size=0.13)
```

